---
title: 'Linear regression: defense ~ .'
subtitle: '2021--2022'
author: 'fnaufel, romulor3'
email: 'https://fnaufel.github.io/'
date: '   (v. `r format(Sys.Date(), "%d/%m/%Y")`)'
lang: 'en'

output: 
  # To install these output formats, run
  #   install.packages("devtools")
  #   devtools::install_github("fnaufel/fnaufelRmd")
  fnaufelRmd::html_report:
    number_sections: no
  fnaufelRmd::pdf_report:
    number_sections: no

# LaTeX / pdf options
#
# For more options, see
# https://pandoc.org/MANUAL.html#variables-for-latex
documentclass: article
classoption: '11pt'
geometry: 'margin=1in'
bibliography: []
biblio-style: apalike
link-citations: yes
---

```{r setup, include=FALSE}
# The next command configures MANY things and loads quite a few packages.
# 
# If you want to see what's being done, execute 
# 
#   cat(
#     system.file(
#       "rmarkdown/resources/R/_common_report.R", 
#       package = "fnaufelRmd"
#     )
#   )
# 
# to find out the location of the file. Then open the file.
# 
# If you want to change the configuration, copy the file, edit it, and
# source it instead of the package file. 
# 
# Or simply write your commands here in this code chunk.

source(
  system.file(
    "rmarkdown/resources/R/_common_report.R",
    package = "fnaufelRmd"
  )
)

# Summarytools options
st_options(
  lang = 'en',
)


library(janitor)
```


# Introduction

???


# Read the data

## Per game

```{r cache=TRUE}
df_per_game <- read_csv(
  './data/2022-12-27-per-player-per-game.csv'
) %>% 
  clean_names()
```

```{r}
df_per_game %>% glimpse()
```

## Advanced

```{r cache=TRUE}
df_advanced <- read_csv(
  './data/2022-12-27-per-player-advanced.csv'
) %>% 
  clean_names()
```

```{r}
df_advanced %>% glimpse()
```

* Empty columns `x20` and `x25`:

  ```{r}
  df_advanced %>% 
    count(x20, x25)
  ```

  Delete them:
  
  ```{r}
  df_advanced <- df_advanced %>% 
    select(-x20, -x25)
  ```
  
  
# Data dictionary

## Per game

- **Rk:** Rank.

- **Player:** Player name.

- **Pos:** Position.

- **Age:** Player's age on February 1 of the season.

- **Tm:** Team.

- **G:** Games.

- **GS:** Games Started.

- **MP:** Minutes Played Per Game.

- **FG:** Field Goals Per Game.

- **FGA:** Field Goal Attempts Per Game.

- **FG%:** Field Goal Percentage.

- **3P:** 3-Point Field Goals Per Game.

- **3PA:** 3-Point Field Goal Attempts Per Game.

- **3P%:** 3-Point Field Goal Percentage.

- **2P:** 2-Point Field Goals Per Game.

- **2PA:** 2-Point Field Goal Attempts Per Game.

- **2P%:** 2-Point Field Goal Percentage.

- **eFG%:** Effective Field Goal Percentage. Adjusts for a 3-point field goal being worth one more point than a 2-point field goal.

- **FT:** Free Throws Per Game.

- **FTA:** Free Throw Attempts Per Game.

- **FT%:** Free Throw Percentage.

- **ORB:** Offensive Rebounds Per Game.

- **DRB:** Defensive Rebounds Per Game.

- **TRB:** Total Rebounds Per Game.

- **AST:** Assists Per Game.

- **STL:** Steals Per Game.

- **BLK:** Blocks Per Game.

- **TOV:** Turnovers Per Game.

- **PF:** Personal Fouls Per Game.

- **PTS:** Points Per Game.

- **Player-additional:** Unique identifier.


## Advanced

- **Rk:** Rank.

- **Player:** Player name.

- **Pos:** Position.

- **Age:** Player's age on February 1 of the season.

- **Tm:** Team.

- **G:** Games.

- **MP:** Minutes Played.

- **PER:** Player Efficiency Rating. A measure of per-minute production standardized such that the league average is 15.

- **TS%:** True Shooting Percentage. A measure of shooting efficiency that takes into account 2-point field goals, 3-point field goals, and free throws.

- **3PAr:** 3-Point Attempt Rate. Percentage of FG Attempts from 3-Point Range

- **FTr:** Free Throw Attempt Rate. Number of FT Attempts Per FG Attempt

- **ORB%:** Offensive Rebound Percentage. An estimate of the percentage of available offensive rebounds a player grabbed while they were on the floor.

- **DRB%:** Defensive Rebound Percentage. An estimate of the percentage of available defensive rebounds a player grabbed while they were on the floor.

- **TRB%:** Total Rebound Percentage. An estimate of the percentage of available rebounds a player grabbed while they were on the floor.

- **AST%:** Assist Percentage. An estimate of the percentage of teammate field goals a player assisted while they were on the floor.

- **STL%:** Steal Percentage. An estimate of the percentage of opponent possessions that end with a steal by the player while they were on the floor.

- **BLK%:** Block Percentage. An estimate of the percentage of opponent two-point field goal attempts blocked by the player while they were on the floor.

- **TOV%:** Turnover Percentage. An estimate of turnovers committed per 100 plays.

- **USG%:** Usage Percentage. An estimate of the percentage of team plays used by a player while they were on the floor.

- **OWS:** Offensive Win Shares. An estimate of the number of wins contributed by a player due to offense.

- **DWS:** Defensive Win Shares. An estimate of the number of wins contributed by a player due to defense.

- **WS:** Win Shares. An estimate of the number of wins contributed by a player.

- **WS/48:** Win Shares Per 48 Minutes. An estimate of the number of wins contributed by a player per 48 minutes (league average is approximately.100)

- **OBPM:** Offensive Box Plus/Minus. A box score estimate of the offensive points per 100 possessions a player contributed above a league-average player, translated to an average team.

- **DBPM:** Defensive Box Plus/Minus. A box score estimate of the defensive points per 100 possessions a player contributed above a league-average player, translated to an average team.

- **BPM:** Box Plus/Minus. A box score estimate of the points per 100 possessions a player contributed above a league-average player, translated to an average team.

- **VORP:** Value over Replacement Player. A box score estimate of the points per 100 TEAM possessions that a player contributed above a replacement-level (-2.0) player, translated to an average team and prorated to an 82-game season. Multiply by 2.70 to convert to wins over replacement.

- **Player-additional:** Unique identifier.


# Cleaning the data

## Per game

* Delete `rk` column:

    ```{r}
    df_per_game <- 
      df_per_game %>% 
        select(-rk)
    ```

* Rename id column:

    ```{r}
    df_per_game <- 
      df_per_game %>% 
        rename(player_id = player_additional)
    ```

* Find players that appear more than once and keep only the row that has the totals:

    ```{r}
    dups <- df_per_game %>% 
      count(player_id, sort = TRUE) %>% 
      filter(n > 1)
    ```
    
  ```{r}
  dups
  ```

  ```{r}
  df_per_game %>% 
    filter(player_id %in% dups$player_id) %>% 
    select(player_id, player, tm)
  ```

    ```{r}
    df_per_game <- 
      df_per_game %>% 
        filter(
          !(player_id %in% dups$player_id) |
          tm == 'TOT'
        )
    ```

* Rename all columns

    ```{r}
    original_names <- names(df_per_game)
    new_names <- c(
      'player',
      'position',
      'age',
      'team',
      'games',
      'games_started',
      'minutes_played',
      'goals_scored',
      'goal_attempts',
      'goal_pct',
      'goals_scored_3p',
      'goal_attempts_3p',
      'goal_pct_3p',
      'goals_scored_2p',
      'goal_attempts_2p',
      'goal_pct_2p',
      'goals_effective_pct',
      'free_throws_scored',
      'free_throw_attempts',
      'free_throw_pct',
      'rebounds_offense',
      'rebounds_defense',
      'rebounds_total',
      'assists',
      'steals',
      'blocks',
      'turnovers',
      'fouls',
      'points_scored',
      'player_id'
    )
    
    names(new_names) <- original_names
    
    paste(
      names(new_names), 
      new_names, 
      sep = ' -> ', 
      collapse = '\n'
    ) %>% 
      cat()
    ```

    ```{r}
    df_per_game <- df_per_game %>% 
      rename_with(
        function(x) { new_names[x] }
      )
    ```

  ```{r}
  df_per_game
  ```

### Per game: summary

```{r}
df_per_game %>% 
  dfSummary(silent = TRUE) %>% 
  print(method = 'render')
```

### Per game: notes

* Approximately $16\%$ of players played for two or more teams during the season.

* On average, a player plays only $19$ minutes per game.

* Columns that should contain percentages actually contain proportions.

* All columns that contain proportions have some missing values. Fortunately, all of them will be discarded before we build the model, as they are derived --- therefore, perfectly correlated with other columns.
